---
title: "Seven species, seven answers: whose traits move which stand"
format:
  html:
    toc: true
    code-fold: true
    fig-width: 9
    fig-height: 6
---

```{r setup, message=FALSE, warning=FALSE}
pkg_root <- if (file.exists("DESCRIPTION")) "." else ".."
if (requireNamespace("pkgload", quietly = TRUE)) {
  pkgload::load_all(pkg_root, quiet = TRUE)
} else {
  library(plant)
}
library(odelia)
library(ggplot2)
library(patchwork)
library(dplyr)
library(tidyr)

source(file.path(pkg_root, "overstorey_staging", "gradient_demo_helpers.R"))
theme_set(theme_bw(base_size = 12))

# Generated by overstorey_staging/gradient_fingerprints.R (about 25 minutes).
F <- readRDS(file.path(pkg_root, "overstorey_staging",
                       "gradient_fingerprints.rds"))
METRIC <- F$metric
ok <- Filter(function(r) !is.null(r$chosen), F$rows)
```

## The question a gradient answers per species

Parameterise a stand at one species, take one reverse sweep, and you have that
species' whole sensitivity vector: how its stand's above-ground mass responds to
every parameter at once. Do it for seven species and the rows are named organisms
rather than abstract directions, so the comparison is one a field ecologist can
argue with — *do these two respond to traits the same way, and which one is
dynamically distinctive?*

The seven span the south-eastern Australian wet sclerophyll to
cool-temperate-rainforest gradient, plus the dry sclerophyll end.

```{r species-table}
F$species |>
  transmute(Species = species, Common = common, Niche = niche,
            `LMA (kg/m²)` = lma, `Wood density (kg/m³)` = rho,
            `Height at maturation (m)` = hmat) |>
  knitr::kable(caption = paste(
    "Representative trait values assembled for this demonstration.",
    "A study takes them from AusTraits; do not cite these."))
```

`hmat` is height at **maturation**, not maximum height — a mountain ash first
reproduces near 16 m, not 90.

## Two things this demo cannot do, and why

**Seed mass is held at the TF24 default for every species.** That is a finding,
not a tidy-up. `fecundity_dt` divides by `(omega + a_f3)` and hyperpar sets
`a_f3 = 3·omega`, so offspring production goes as `1/omega`. TF24 is calibrated
at `omega` = 3.8e-5 kg, and real Australian seeds run one to two orders below it.
Run with their own seed masses, six of these seven species produced gradients
between 1e+05 and 1e+128 or would not run — **ordered exactly by how far below the
default they sat**, with the sole well-behaved species the only one whose seed is
*heavier* than the default.

⚠️ **DO NOT parameterise TF24 from a trait database without checking seed mass
against 3.8e-5 kg.** The failure is silent at the top of the range — a plausible
seed and a gradient of 1e+58 — and only becomes a hard error two orders down.

**And a single gradient per species is not trustworthy.** The maturation switch
`a_f1/(1+exp(a_f2(1−h/hmat)))` is nearly a step at `hmat`, a cohort crosses it
inside one or two ODE steps, and which step catches the crossing sets the
gradient and flips under changes of 1e-4 in a trait. So each species is evaluated
at a ladder of tiny `lma` offsets and an answer is kept only where **three
consecutive rungs agree** within a factor of `r F$agree`.

## The ladders, which are the honest part

```{r ladders, fig.height=5}
lad <- bind_rows(lapply(F$rows, function(r) tibble::tibble(
  common = r$common, rung = seq_along(r$rungs), offset = r$rungs,
  norm = r$norms, picked = !is.na(r$pick) & seq_along(r$rungs) == r$pick)))

ggplot(lad, aes(offset, norm)) +
  geom_line(colour = "grey65") +
  geom_point(aes(colour = picked), size = 2) +
  geom_point(data = subset(lad, picked), shape = 21, size = 4,
             colour = "#1b7837", stroke = 1.1, fill = NA) +
  facet_wrap(~common, scales = "free_y", ncol = 4) +
  scale_y_continuous(trans = "log10", labels = scales::label_scientific()) +
  scale_colour_manual(values = c(`FALSE` = "#b2182b", `TRUE` = "#1b7837"),
                      guide = "none") +
  labs(title = "Nudging LMA by fractions of a percent, per species",
       subtitle = paste("Ringed = the kept rung, the middle of three that",
                        "agree.\nSpikes are unresolved maturation crossings,",
                        "not ecology."),
       x = "relative offset applied to LMA", y = "elasticity norm (log scale)")
```

Two species are smooth across the whole ladder; the rest have at least one rung
orders of magnitude out — old man banksia opens at 8.3e+05 before settling near
3.6. **The spikes are not a property of the species**: they are the maturation
crossing landing inside a different ODE step. A demo that took one gradient per
species and plotted it would have published those spikes as ecology.

```{r survivors}
cat(sprintf("%d of %d species reached three consecutive agreeing rungs\n",
            length(ok), length(F$rows)))
for (r in F$rows) if (is.null(r$chosen))
  cat("  no stable run of three:", r$species, "\n")

# Passing the test is not the same as having converged. The ladder's own range
# does not separate the two -- a species with a clean plateau still has a wide
# range if it spikes anywhere else -- so the test is against the other species:
# a kept value far above the cohort has not settled onto the same kind of answer.
kept <- vapply(ok, function(r) {
  e <- r$chosen$elasticity[METRIC, ]; sqrt(sum(e[is.finite(e)]^2)) }, numeric(1))
names(kept) <- vapply(ok, `[[`, character(1), "common")
for (n in names(kept))
  if (kept[[n]] > 4 * stats::median(kept))
    cat(sprintf("  kept value %.1f against a cohort median of %.1f, not settled: %s\n",
                kept[[n]], stats::median(kept), n))
```

⚠️ **Hazel pomaderris satisfies the letter of the test and not its spirit.** Its
ladder falls steadily from 66 to 14 and then spikes to 698; the kept triple sits
on that slope rather than on a plateau, and its elasticities below are an order of
magnitude larger than every other species'. Read that row as "not resolved",
not as "extremely sensitive".

## The fingerprints

```{r fingerprint, fig.height=6}
el <- lapply(ok, function(r) r$chosen$elasticity[METRIC, ])
names(el) <- vapply(ok, `[[`, character(1), "common")

# The traits a database reports, plus the architecture parameters that set how
# leaf area converts to height. a_f1 is deliberately excluded from the ranking:
# it is the amplitude of the maturation switch, defaults to the boundary value
# 1.0, and no database reports it -- so it wins on wood by restating the
# allocation equation rather than by saying anything ecological.
show <- c("lma", "rho", "hmat", "a_l1", "a_l2", "theta", "vcmax_25",
          "curv_fact_colim", "k_l", "r_l")
d <- bind_rows(lapply(names(el), function(n) {
  e <- el[[n]]
  tibble::tibble(common = n, param = sub("^1\\.", "", names(e)),
                 e = as.numeric(e))
})) |> filter(param %in% show)

ggplot(d, aes(param, common, fill = e)) +
  geom_tile(colour = "white") +
  geom_text(aes(label = sprintf("%+.1f", e)), size = 2.9) +
  scale_fill_gradient2(low = "#b2182b", mid = "white", high = "#2166ac",
                       midpoint = 0, limits = c(-5, 5), oob = scales::squish) +
  labs(title = "Elasticity of stand above-ground mass, by species",
       subtitle = "Percent change in mass per percent change in the parameter; colour saturates beyond ±5",
       x = NULL, y = NULL, fill = "%/%") +
  theme(axis.text.x = element_text(angle = 40, hjust = 1),
        panel.grid = element_blank())
```

```{r levers}
bind_rows(lapply(names(el), function(n) {
  e <- el[[n]][paste0("1.", show)]
  o <- order(abs(e), decreasing = TRUE)[1:3]
  tibble::tibble(Species = n,
                 `Largest lever` = sub("^1\\.", "", names(e)[o[1]]),
                 `%/%` = round(as.numeric(e[o[1]]), 2),
                 Runner_up = sub("^1\\.", "", names(e)[o[2]]),
                 Third = sub("^1\\.", "", names(e)[o[3]]))
})) |> knitr::kable(caption = "Among the traits a database reports and the architecture parameters.")
```

## Who resembles whom

```{r angles, fig.height=5}
ns <- names(el)
A <- outer(seq_along(ns), seq_along(ns),
           Vectorize(function(i, j) gd_angle(el[[i]], el[[j]])))
dimnames(A) <- list(ns, ns)

as.data.frame(as.table(A)) |>
  setNames(c("a", "b", "deg")) |>
  ggplot(aes(a, b, fill = deg)) +
  geom_tile(colour = "white") +
  geom_text(aes(label = sprintf("%.0f", deg)), size = 3) +
  scale_fill_viridis_c(option = "cividis", direction = -1) +
  labs(title = "Angle between species' sensitivity vectors",
       subtitle = "0° = responds to traits identically; 90° = responds to unrelated things",
       x = NULL, y = NULL, fill = "deg") +
  theme(axis.text.x = element_text(angle = 40, hjust = 1),
        panel.grid = element_blank())
```

An angle near zero means two species' stands respond to the trait space the same
way, whatever their standing biomass. A large angle means they are dynamically
distinct — the same trait perturbation moves them differently, or in opposite
directions. That is a claim about *dynamics* rather than about state, and it is
not visible in any single-run output.

### What the heatmap actually shows

Three bands, and the middle one is the interesting one.

**Shared.** `rho` and `theta` sit near +0.8 for every species, and `k_l` and
`vcmax_25` near zero for every species. Those columns carry no information about
who the species is; they are how this model responds to those parameters, full
stop.

**Divergent, and by sign.** `a_l1` and `hmat` — the leaf-area-to-height allometry
and the maturation height — **change sign between species**. Silver wattle and
old man banksia gain mass when `a_l1` rises and lose it when `hmat` rises; myrtle
beech and mountain ash do the opposite, and by more. Two species can therefore
sit at similar standing biomass and want opposite things from the same trait,
which is precisely the distinction a fingerprint exists to find and a mean
response would hide.

**Unresolved.** Hazel pomaderris, an order of magnitude above the rest, for the
reason given above. It is left in the figure rather than dropped, because a demo
that quietly removes its awkward row teaches the wrong lesson.

A caution on mechanism: the sign split does not line up cleanly with `hmat`
alone — black she-oak and old man banksia share `hmat` = 5 and disagree — so it
is the trait combination, not maturation height by itself. Naming the mechanism
would need the trajectories, not the gradients.

## What it cost

`r length(ok)` species with usable answers, `r length(F$rungs)` rungs each on a
`r F$lifetime`-year stand sharing one `r F$nodes`-node schedule. Each gradient
gives all 47 columns for about three forward runs; differencing for them would
have cost 47 runs apiece, and the multiple grows with the number of species in
the stand rather than staying fixed.

## Reading this honestly

**The ladder is not optional.** Rerun any of this without it and some rows will be
needles reported as ecology.

**Seed mass is missing from the comparison**, and it is one of the four traits a
database actually reports. Until TF24 is calibrated over realistic seed masses,
these fingerprints describe leaf economics, wood density and maturation height
only.

**One species, one stand.** Each row is a monoculture, not a member of a mixed
community. The cross-species question — how species *i* responds to species *j*'s
traits — needs a per-species census metric, which the model does not currently
carry: the three metrics are stand totals.
