---
title: "A phase diagram of trait space: where the model changes its mind"
format:
  html:
    toc: true
    code-fold: true
    fig-width: 9
    fig-height: 6
---

```{r setup, message=FALSE, warning=FALSE}
pkg_root <- if (file.exists("DESCRIPTION")) "." else ".."
if (requireNamespace("pkgload", quietly = TRUE)) {
  pkgload::load_all(pkg_root, quiet = TRUE)
} else {
  library(plant)
}
library(odelia)
library(ggplot2)
library(patchwork)
library(dplyr)
library(tidyr)

source(file.path(pkg_root, "overstorey_staging", "gradient_demo_helpers.R"))
theme_set(theme_bw(base_size = 12))

# Generated by overstorey_staging/gradient_phase_grid.R, which takes about 45
# minutes. Not regenerated here: a document that recomputes an hour of gradients
# is a document nobody renders.
grid_path <- file.path(pkg_root, "overstorey_staging", "gradient_phase_grid.rds")
G <- readRDS(grid_path)
METRIC <- "mass_above_ground"
```

## The idea

One gradient per grid cell, then throw the gradients away and keep three scalars
from each. The scalars are what a map can show:

* **how responsive** the stand is here, `‖e‖` over the elasticities
* **how far its opinion has rotated** from the centre of the grid, the angle
  between this cell's elasticity vector and the middle one
* **which parameter is the largest lever** here — a categorical map, whose
  boundaries are where the ranking flips

The third is the phase diagram proper: a boundary in it is a line across which a
different parameter becomes the thing that matters most. Whether this fixture has
any such boundary is a question rather than a premise, and the answer below is
mostly no — which is itself worth knowing, and is the sort of thing 64 gradients
are for.

```{r frame}
cells <- Filter(function(c) is.null(c$error), G$cells)
centre_i <- which.min(abs(log(vapply(cells, `[[`, numeric(1), "fx"))) +
                      abs(log(vapply(cells, `[[`, numeric(1), "fy"))))
centre_e <- cells[[centre_i]]$elasticity[METRIC, ]

rows <- lapply(cells, function(c) {
  e <- c$elasticity[METRIC, ]
  lev <- gd_largest_lever(c$elasticity, METRIC)
  tibble::tibble(
    fx = c$fx, fy = c$fy,
    lma = c$traits[["lma"]], hmat = c$traits[["hmat"]],
    value = c$value[[METRIC]],
    norm = sqrt(sum(e[is.finite(e)]^2)),
    angle = gd_angle(centre_e, e),
    lever = sub("^1\\.", "", lev$name),
    margin = lev$margin,
    refused = c$refused[[METRIC]])
})
df <- bind_rows(rows)

# The grid is spaced geometrically, so the cells are not the same width in trait
# units. geom_raster assumes they are and leaves gaps between them, so the maps
# are drawn on the grid INDEX and the axes are labelled with the trait values the
# indices stand for.
ux <- sort(unique(df$lma)); uy <- sort(unique(df$hmat))
df$ix <- match(df$lma, ux); df$iy <- match(df$hmat, uy)
axis_x <- scale_x_continuous(breaks = seq_along(ux), labels = signif(ux, 2))
axis_y <- scale_y_continuous(breaks = seq_along(uy), labels = signif(uy, 3))
LAB <- labs(x = "LMA", y = "height at maturity")
# Eight labels on a narrow panel collide, so they are turned.
TURN <- theme(axis.text.x = element_text(angle = 45, hjust = 1))

cat(sprintf("%d of %d cells answered; %d nodes shared by all of them; %s axes\n",
            nrow(df), length(G$cells), G$nodes, paste(G$axes, collapse = " x ")))
```

## The stand itself, for reference

Before the gradients: the metric they are gradients of. A phase boundary that
coincides with a fold in this surface is a different claim from one that crosses
open ground.

```{r value, fig.height=4.6}
ggplot(df, aes(ix, iy, fill = value)) +
  geom_raster() + axis_x + axis_y + LAB + TURN +
  scale_fill_viridis_c(option = "mako") +
  labs(title = "Above-ground mass across the grid", fill = "mass")
```

## Responsiveness and rotation

```{r norm-angle, fig.height=4.6}
# Four cells are two orders of magnitude above the rest (see "The cells that do
# not belong", below). On a scale they set, every other cell is the same colour
# and the map says nothing -- so the scale is clipped at the 90th percentile and
# the cells above it are marked instead of coloured.
clip <- stats::quantile(df$norm, 0.90)
p1 <- ggplot(df, aes(ix, iy, fill = pmin(norm, clip))) +
  geom_raster() + axis_x + axis_y + LAB + TURN +
  geom_point(data = subset(df, norm > clip), shape = 21, size = 3,
             fill = NA, colour = "white", stroke = 1.1) +
  scale_fill_viridis_c(option = "rocket") +
  labs(title = "Responsiveness",
       subtitle = "norm of the elasticity vector; ringed cells exceed the scale",
       fill = "|e|")

p2 <- ggplot(df, aes(ix, iy, fill = angle)) +
  geom_raster() + axis_x + axis_y + LAB + TURN +
  scale_fill_viridis_c(option = "cividis") +
  labs(title = "Rotation from the grid centre",
       subtitle = "angle between elasticity vectors, degrees",
       fill = "deg", y = NULL)

p1 + p2
```

The angle map is the more interesting of the two. Responsiveness says how hard
the stand reacts; rotation says whether it reacts to *different things*. A region
of constant colour in the right panel is a region where one linearisation
describes the whole neighbourhood — which is the licence for reasoning about the
model locally. Most of this grid is such a region: the background sits under
about 20°, and the exceptions are isolated single cells rather than any
boundary.

## The phase diagram

```{r phase, fig.height=5.4}
lev_n <- df |> count(lever, sort = TRUE)
keep <- lev_n$lever[lev_n$n >= 2]
dfp <- df |> mutate(lever2 = ifelse(lever %in% keep, lever, "other"))

ggplot(dfp, aes(ix, iy, fill = lever2)) +
  geom_raster() + axis_x + axis_y + LAB + TURN +
  # Cells whose ranking is not decisive are drawn hollow, because a colour there
  # states a fact the numbers do not support.
  geom_point(data = subset(dfp, margin < 1.2), shape = 4, size = 2.4,
             colour = "grey15") +
  geom_text(aes(label = sprintf("%.1f", margin)), size = 2.5,
            colour = "grey15", vjust = -1.1) +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "Which parameter is the largest lever here",
       subtitle = paste("Number = times larger than the runner-up.",
                        "A cross marks a cell below 1.2, where the ranking is",
                        "not decisive."),
       fill = NULL)
```

**This grid has one phase, not several, and that is the result.** Over a ±45%
window in both traits, `a_f1` is the largest lever on above-ground mass almost
everywhere. The scattered cells where something else wins are not a region: they
are isolated, and they sit where the margin is low, which is where the ranking is
a near-tie rather than a different regime.

⚠️ **`a_f1` winning is not an ecological finding, and a demo should not sell it as
one.** It is the ceiling on the fraction of production a mature plant sends to
reproduction, and it defaults to 1.0 — meaning a plant past `hmat` puts
everything into seed and nothing into wood. That it is the largest lever on
*wood* restates the allocation equation. Three further reasons to distrust it as
a headline: no trait database reports it, so there is nothing to compare against;
its default sits exactly on the boundary of the admissible set, since a fraction
cannot exceed one, so half of its perturbation direction is meaningless; and it
is the amplitude of the very switch that produces the needles below. **Rank among
the traits a database reports if the question is ecological.**

**Read the margins, not just the colours.** A cell whose margin is 1.05 has two
levers within five percent of each other, and which one wins there is not a fact
about the model. Here only one cell of `r nrow(df)` falls below 1.2, so the
ranking is decisive nearly everywhere — and what it decides is the same thing
nearly everywhere.

What *does* have structure is the margin itself. It falls from about 4.7 at short
maturity heights to about 1.7 at tall ones: `a_f1` stays the largest lever, and
stops being overwhelmingly so. A single dominant lever weakening smoothly across
trait space is a more useful thing to know than a speckled map of ties, and it is
the honest reading of this fixture.

A note on scope, since a negative result invites over-reading: this is one
species, one 30-year stand, one metric, and two of the 47 parameters varied. A
grid over a different pair, or a metric other than mass, may well be partitioned.

```{r margin, fig.height=4}
ggplot(df, aes(margin)) +
  geom_histogram(bins = 24, fill = "#3b6ea5") +
  geom_vline(xintercept = 1.2, linetype = 2, colour = "#b2182b") +
  scale_x_continuous(trans = "log10") +
  labs(title = "How decisive is the ranking, cell by cell",
       subtitle = "left of the line, the largest lever is not meaningfully largest",
       x = "ratio of largest to runner-up lever", y = "cells")
```

## The cells that do not belong

Four cells carry gradients two orders of magnitude above their neighbours, and
the same cells are anomalous in all three summaries above. A quantity that jumps
100× in one cell and returns to baseline in the next is not describing a smooth
trait surface, so it is worth saying what these are — and what they are not.

```{r outliers, fig.height=4}
q <- stats::quantile(df$norm, c(0.5, 0.9, 1))
cat(sprintf("|e| median %.2f, 90th %.2f, max %.2f -- max is %.0fx the median\n",
            q[1], q[2], q[3], q[3] / q[1]))

ggplot(df, aes(norm)) +
  geom_histogram(bins = 30, fill = "#3b6ea5") +
  scale_x_continuous(trans = "log10") +
  labs(title = "Responsiveness is not a smooth field across this grid",
       subtitle = "a tight body of cells, and a handful two decades away",
       x = "norm of the elasticity vector (log scale)", y = "cells")
```

Two explanations were tested against the fixture and **both are wrong**:

* *An elasticity dividing by a near-zero metric.* It is not: the census at those
  cells is entirely ordinary — 8.4, 10.1, 10.0, 13.8 against a grid median of
  10.9, a two-fold spread over the whole grid — and the **raw** gradient at the
  worst cell is 5.0e+07. The spike is in the gradient, not in the normalisation.
* *A profit curvature near its floor.* The interior derivation divides by that
  curvature, so a small one would inflate the gradient without refusing. It is
  not that either: the worst cell still answers with
  `gradient_curvature_floor` raised a hundredfold, to 1e-1.

### What they actually are

Sweeping `hmat` through the worst cell, with `lma` held at its value, finds not
one sharp feature but several needles at irregular spacing, with entirely
ordinary values between them — while the mass runs smoothly and monotonically
through the lot.

```{r needle, fig.height=4.2}
# Measured by sweeping hmat through the worst cell of this grid; see
# docs/design/measurements.md, "The gradient's needles, and what they are".
needle <- tibble::tribble(
  ~d,      ~norm,    ~mass,
  -3e-2,     7.78,  8.0772,
  -1e-2,     6.73,  8.3037,
  -3e-3,     6.56,  8.3956,
  -1e-3,     7.04,  8.4204,
  -3e-4,     6.69,  8.4286,
   0,     1276.73,  8.4322,
   3e-4,    11.97,  8.4356,
   1e-3,  1888.61,  8.4435,
   3e-3,     6.28,  8.4641,
   1e-2,   193.93,  8.5344,
   3e-2,     6.30,  8.7640)

p1 <- ggplot(needle, aes(d, norm)) +
  geom_line(colour = "grey60") + geom_point(size = 2, colour = "#b2182b") +
  scale_y_continuous(trans = "log10") +
  labs(title = "The derivative", x = "relative offset in hmat", y = "|e|")
p2 <- ggplot(needle, aes(d, mass)) +
  geom_line(colour = "grey60") + geom_point(size = 2, colour = "#2166ac") +
  labs(title = "The quantity it is a derivative of", x = "relative offset in hmat",
       y = "above-ground mass")
p1 + p2 + patchwork::plot_annotation(
  title = "Needles in the derivative over a smooth value",
  subtitle = "A feature narrower than 3e-4 in a parameter is not ecology")
```

The cause is the maturation switch, and it is the same `a_f1` the map above is
coloured by:

$$\text{fraction to reproduction}(h) = \frac{a_{f1}}{1 + \exp\!\big(a_{f2}(1 - h/h_{mat})\big)}$$

With `a_f2` = 50 that is very nearly a **step** at `hmat` — 0.7% of `a_f1` at
0.9·hmat, 99.3% at 1.1·hmat. A cohort grows through the whole transition inside
one or two ODE steps. The reverse sweep differentiates *the discrete trajectory
that was actually taken*, so the gradient is dominated by which step happened to
catch the crossing, and that changes discontinuously with `hmat`. The integrated
census averages over the transition and never notices.

Softening the step confirms it, against a calm cell as the control:

| `a_f2` | ‖e‖ at the needle | `e[a_f1]` | ‖e‖ at the control |
|---|---|---|---|
| 50 (default) | **1276.73** | +848.67 | 5.95 |
| 25 | 7.83 | −7.31 | |
| 10 | 12.45 | −10.84 | 10.04 |
| 5 | 8.55 | −7.61 | |

Halving `a_f2` collapses the needle 163× and restores the negative sign every
calm cell has. Softening does not create pathology at the control, so it is the
sharpness that matters, not the change.

**So the gradient is not wrong.** It is the exact derivative of a trajectory that
resolves the maturation switch crudely. The principled repair is event detection —
force a step at `h = hmat` so the crossing is resolved the same way every time —
and nothing here needs it yet.

⚠️ **DO NOT average over the needles, and DO NOT form a covariance from a grid
containing one.** A single cell 211× the median dominates any second-moment
summary, which is the first thing an active-subspace workflow computes. Treat a
gradient far outside its neighbourhood as a *diagnostic*: it marks an unresolved
event. The cheap response is to nudge the trait a fraction of a percent and keep
the answer that two adjacent nudges agree on.

## Why every cell shares one node schedule

The grid is generated with **one** node-introduction schedule, refined once at the
centre and transplanted to every cell. That is a correctness requirement, not an
economy.

`build_schedule` inserts a node wherever removing one would move the forward
leaf-area or seed-rain integral by more than `schedule_eps`. It is a criterion
about the forward integrals, and the gradient is far more sensitive to the
resulting node set than the census is. Measured at one point, refining at
`schedule_eps` of 0.02, 0.01, 0.005 and 0.002:

| `schedule_eps` | nodes | census, vs finest | gradient, vs finest |
|---|---|---|---|
| 0.02 | 112 | 3.8e-04 | 1.2e-02 |
| 0.01 | 118 | 8.4e-04 | 9.7e-03 |
| 0.005 | 125 | 5.5e-04 | 1.20 |
| 0.002 | 153 | — | — |

The census is settled to four decimal places throughout. The gradient is not
settled at all, and is not even monotone. What it is sensitive to is *which node
set it got*: hold the node set fixed and the gradient turns smoothly — stepping
LMA by 2% five times moves the elasticity vector by 1.4°, 0.5°, 0.8°, 0.5°, 1.5°
— and the same five points refined individually give bit-identical answers,
because refinement happens to choose the same 112 nodes.

So the gradient is smooth wherever the node set is constant and jumps where it
changes. Sharing one schedule makes the node set constant by construction, and
every boundary in the map above is then the model changing its mind rather than
the mesh changing underneath it.

⚠️ **Do not also transplant `ode_times`.** Pinning the ODE step grid chosen at the
centre makes the leaf fail to place an operating point out at the corners, and
the gradient refuses every metric on about a third of them. The node set is
shared; the stepper stays free.

## What this cost

`r nrow(df)` gradients at `r G$nodes` nodes on a `r G$lifetime`-year stand, one
schedule refinement, about 45 minutes. Differencing for the same 47 columns would
have been roughly 16 times that, and the multiple grows with the number of
species in the stand.
